Two subtle problems with over-representation analysis

### 2024-09-17 manuscript/manuscript.Rmd
more corrections

According to a study that used simulated differential expression data, FCS accuracy was
superior to ORA over a range of experimental designs [@Kaspi2020-rn].
A study of 1366 PubMed Central articles featuring enrichment analysis published in 2019
indicates that seven of the top eight most popular tools were based on ORA (according to the
supplementary data)[@Wijesooriya2022-fw].
The discrepancy between performance and popularity is likely due to the relative ease of
conducting ORA as compared to FCS.
A minimal ORA might involve pasting a list of gene symbols into a text box on a website,
with results appearing nearly instantly.
On the other hand, FCS typically involves installing specific software and dealing with a
dataset representing every detected gene (~20,000 rows) and ensuring that the file
format from upstream tools is compatible with FCS packages.

@@

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets, each containing 30 genes, was generated.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to clusterProfiler which suffers from both problems,
clusterProfiler with the background problem fix, clusterProfiler with the FDR problem fix,
clusterProfiler with the fixes to background and FDR problems, an ORA function called `fora`
from `fgsea` bioconductor package v1.30.0 that does not suffer these problems, and `fgsea` a
fast implementation of the preranked gene set enrichment analysis (GSEA) algorithm
[@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest
p-values were used for ORA, like a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.
All analyses were undertaken with R v4.4.1 in a Docker container.

@@

To understand the sensitivity and false positive rate of ORA and FCS methods, we employed a downsampling
approach to a previously published dataset derived from 37 lung cancer patients, having tumour and normal
tissues for each patient [@Cho2022-ri].
These RNA-seq based gene expression counts were downloaded from NCBI GEO under accession number GSE158420.
After reading them into R, we noticed some gene names were converted to dates [@Abeysooriya2021-qt], so we
used the HGNChelper package to fix them [@Oh2022-gj].
DESeq2 was then used for differential expression analysis correcting for patient-of-origin to identify genes
differentially expressed in tumour compared to normal tissue.
FCS was conducted using fgsea using the DESeq2 test statistic to score each gene.
ORA was conducted using fora, using genes with differential expression FDR<0.05.
Separate tests were conducted for up- and down-regulated genes.
For the ORA background, genes with a mean of 10 or more reads per sample on average across the comparison were
included.
Reactome pathways were used.
For downsampling, sample size was varied between 2 and 30 patients, selected pseudorandomly with a set seed,
followed by parallel analysis with fgsea and fora pipelines, using a significance filter of FDR<0.05.
We then calculated the number of pathways identified in downsampled data that were consistent with the full
dataset.
The number of consistent pathways gives an indication of sensitivity of each method.
Then, we calculated the number of pathways identified in the downsampled data that were inconsistent with the
full dataset.
The proportion of significant pathways classified as inconsistent provides an estimate of the real false
discovery rate.
This was repeated 50 times at each sample size, and the median values were reported.

@@

A workaround to eliminate the background problem was developed and we used it to compare the
results of ORA of seven datasets with the original and corrected methods for nine different
gene set libraries.
The number of statistically significant results was calculated for each of the gene set
libraries before and after correcting the problem (**Figure 1A**).
Each study was affected to a similar degree (mean uplift 125% ±199%), except for #5 which saw the number of
significant GO terms increase from 8 to 54 after correction (**Table S1**).
The similarity between results, as quantified with the Jaccard similarity, was highest for
Cellmarkers and GTRD at 0.76 and 0.77 respectively, and the lowest similarity was recorded for
HPO and KEGG with 0.20 and 0.48 respectively.
Observed Jaccard scores correlated with the number of genes annotated to one or more functional
categories in the gene set library (**Figure 1B**).
This result suggests that ORA with smaller gene set libraries like HPO and KEGG are impacted
more severely.
We investigated discrepant results for data set #1 (SRP128998) downregulated genes in HPO gene sets.
A scatterplot of significance scores (**Figure 1C**) shows severe deviation in p-values in gene sets in the largest
quintile (109-1382 members detected), suggesting larger gene sets are more severely affected by the background
problem.

@@

The purpose of this work was to thoroughly characterise two subtle problems with ORA, but there
are some bigger problems noted.
Firstly, all variations of ORA were less accurate (based on F1 score) as compared to FCS in this
simulation work, confirming a previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
The higher recall of FCS demonstrated in simulated differential expression profiles and confirmed
with downsampled real cancer RNA-seq data means that researchers will have more power to identify
pathways relevant to the biological processes being studied.
Secondly, many popular ORA tools listed in Table 2 do not provide enrichment scores in their
default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance values when interpreting enrichment results.
Lastly, ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.

>>>>

According to a study that used simulated differential expression data, FCS accuracy was
superior to ORA over a range of experimental designs [@Kaspi2020-rn].
A study of 1366 PubMed Central articles featuring enrichment analysis published in 2019
indicates that seven of the top eight most popular tools were based on ORA (according to the
supplementary data)[@Wijesooriya2022-fw].
The discrepancy between performance and popularity is likely due to the relative ease of
conducting ORA as compared to FCS.
A minimal ORA involves pasting a list of gene symbols into a text box on a website,
with results appearing nearly instantly.
On the other hand, FCS typically involves installing specific software and dealing with a
dataset representing every detected gene (~20,000 rows) and ensuring that the file
format from upstream tools is compatible with FCS packages.

@@

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets, each containing 30 genes, was generated.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0 and 0.6, this value selected randomly for each gene.
Three "control" and "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to clusterProfiler which suffers from both problems,
clusterProfiler with the background problem fix, clusterProfiler with the FDR problem fix,
clusterProfiler with the fixes to background and FDR problems, an ORA function called `fora`
from `fgsea` bioconductor package v1.30.0 that does not suffer these problems, and `fgsea` a
fast implementation of the preranked gene set enrichment analysis (GSEA) algorithm
[@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest
p-values were used for ORA, like a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.
All analyses were undertaken with R v4.4.1 in a Docker container.

@@

To understand the sensitivity and false positive rate of ORA and FCS methods, we employed a downsampling
approach to a previously published dataset derived from 37 lung cancer patients, having tumour and normal
tissues for each patient [@Cho2022-ri].
These RNA-seq based gene expression counts were downloaded from NCBI GEO under accession number GSE158420.
After reading them into R, we noticed some gene names were converted to dates [@Abeysooriya2021-qt], so we
used the HGNChelper package to fix them [@Oh2022-gj].
DESeq2 was then used for differential expression analysis correcting for patient-of-origin to identify genes
differentially expressed in tumour compared to normal tissue.
FCS was conducted using fgsea using the DESeq2 test statistic to score each gene.
ORA was conducted using fora, using genes with differential expression FDR<0.05.
Separate tests were conducted for up- and down-regulated genes.
For the ORA background, genes with a mean of 10 or more reads per sample on average across the comparison were
included.
Reactome pathways were used.
For downsampling, sample size was varied between 2 and 30 patients, selected pseudorandomly with a set seed,
followed by parallel analysis with fgsea and fora pipelines, using a significance filter of FDR<0.05.
We then calculated the number of pathways identified in downsampled data that were consistent with the full
dataset.
The number of consistent pathways gives an indication of sensitivity of each method.
Then, we calculated the number of pathways identified in the downsampled data that were inconsistent with the
full dataset.
The proportion of significant pathways classified as inconsistent provides an estimate of the real false
discovery rate.
This was repeated 50 times at each sample size, and the results were presented as a violin plot.

@@

A workaround to eliminate the background problem was developed and we used it to compare the
results of ORA of seven datasets with the original and corrected methods for nine different
gene set libraries.
The number of statistically significant results was calculated for each of the gene set
libraries before and after correcting the problem (**Figure 1A**).
Each study was affected to a similar degree (31-94% increase), except for #5 which saw the number of
significant GO terms increase from 8 to 54 after correction (**Table S1**).
The similarity between results, as quantified with the Jaccard similarity, was highest for
Cellmarkers and GTRD at 0.76 and 0.77 respectively, and the lowest similarity was recorded for
HPO and KEGG with 0.20 and 0.48 respectively.
Observed Jaccard scores correlated with the number of genes annotated to one or more functional
categories in the gene set library (**Figure 1B**).
This result suggests that ORA with smaller gene set libraries like HPO and KEGG are impacted
more severely.
We investigated discrepant results for data set #1 (SRP128998) downregulated genes in HPO gene sets.
A scatterplot of significance scores (**Figure 1C**) shows severe deviation in p-values in gene sets in the largest
quintile (109-1382 members detected), suggesting larger gene sets are more severely affected by the background
problem.

@@

The purpose of this work was to thoroughly characterise two subtle problems with ORA, but there
are some bigger problems noted.
Firstly, all variations of ORA were less accurate (based on F1 score) as compared to FCS in this
simulation work (**Figure 3A**), confirming a previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
The higher recall of FCS demonstrated in simulated differential expression profiles and confirmed
with downsampled real cancer RNA-seq data means that researchers will have more power to identify
pathways relevant to the biological processes being studied.
Secondly, many popular ORA tools listed in Table 2 do not provide enrichment scores in their
default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance values when interpreting enrichment results.
Lastly, ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.
